### 2024-03-27 doc/sections/mav-cut-elimination.tex
[ agda tex ] Streamline and write up the cut elimination proof

\section{Semantic Cut-Elimination}\label{sec:mav-cut-elimination}

\begin{proposition}
  Let $\Chu(\ClosedLowerSet{\textsc{CFMav}}, \ClosedDay{I})$ be the
  MAV-algebra constructed from the Cut-free proof MAV-frame
  (Proposition \ref{prop:cfmav-frame}), where elements are
  positive/negative pairs of $+$-closed lower sets of formulas.

  \bob{describe the interpretation of atoms}

  The interpretation of SMAV formulas in this MAV-algebra satisfies
  the following two properties for every formula $\vP$:
  \begin{enumerate}
  \item $\eta^+(P) \subseteq \sem{\vP}^-$
  \item $\sem{\vP}^+ \subseteq \eta^+(P) \rightblackspoon \ClosedDay{i}$
%  \item $\forall \vR \in \sem{\vP}^+.~(\vR \vParr \vP) \longrightarrow^* I$
  \end{enumerate}
\end{proposition}

\begin{remark}
  These properties are akin to the \emph{reflect} and \emph{reify}
  functions used in Normalisation by Evaluation (NbE). The
  \emph{reflect} property creates a canonical negative proof of every
  formula assuming that formula (corresponding to $\eta$-expansion in
  NbE. The \emph{reify} property extracts a normal-form proof from the
  semantic model.
\end{remark}

\begin{proof}
  By induction on the structure of the formula $\vP$.
  \begin{itemize}
  \item[$\vUnit$] reflexivity
  \item[A+] reflexivity
  \item[A-] reflexivity
  \item[$\vP \vParr \vQ$] $\eta^+$ preserves monoids (FIXME: proposition)
  \item[$\vP \vTens \vQ$] complicated
  \item[$\vP \vWith \vQ$] $\eta^i$ preserves $+$
  \item[$\vP \vPlus \vQ$] interpreted as a meet, so we reason directly
  \item[$\vP \vSeq \vQ$] $\eta^+$ preserves monoids (FIXME: proposition)
  \end{itemize}
\end{proof}

\subsection{The Okada Property}

>>>>

\section{Semantic Cut-Elimination and Proof Normalisation}
\label{sec:mav-cut-elimination}

\newcommand{\ChuEmbed}{\eta^c}
\newcommand{\ClosedLowerEmbed}{\eta^+}
\newcommand{\LowerEmbed}{\eta}

Let $\Chu(\ClosedLowerSet{\textsc{CFMav}}, \ClosedDay{I})$ be the
MAV-algebra constructed from the Cut-free proof MAV-frame (Proposition
\ref{prop:cfmav-frame}), where elements are positive/negative pairs of
$+$-closed lower sets of formulas. We define the valuation of atoms as
$V(\alpha) = \ChuEmbed(\ClosedLowerEmbed(\vDual{\alpha}))$. By Theorem
\ref{thm:soundness}, we have an interpretation of SMAV formulas
$\sem{P}$ such that if $P \longrightarrow^* I$, then $I \leq
sem{P}$. We now prove our main proposition about the MAV-algebra
\textsc{CFMav} that will allow us to derive the admissibility of Cut
and all the other symmetric rules of SMAV.

\begin{proposition}\label{prop:embedding-sem}
  $\sem{P} \leq \lnot (\ChuEmbed(\ClosedLowerEmbed(P)))$
\end{proposition}

\begin{proof}
  By the definition of the Chu construction, this statement breaks
  down into two inclusions between pairs of $+$-closed lower sets:
  \begin{enumerate}
  \item $\ClosedLowerEmbed(P) \subseteq \sem{\vP}^-$
  \item $\sem{\vP}^+ \subseteq \ClosedLowerEmbed(P) \rightblackspoon^+ \ClosedDay{I}$
  \end{enumerate}
  We prove the second assuming the first. It suffices to prove that
  $\sem{\vP}^+ \ClosedDay{\bullet} \ClosedLowerEmbed(P) \leq
  \ClosedDay{I}$, which follows from the first part and the property
  of all Chu-elements that
  $\sem{\vP}^+ \ClosedDay{\bullet} \sem{\vP}^- \leq \ClosedDay{I}$.

  We prove the first part by induction on $P$. In the cases when
  $P = \vUnit$ or $P = \vDual{\alpha}$, we already have
  $\sem{P}^- = \ClosedLowerEmbed(P)$. When $P = \alpha$, we have
  $\sem{\alpha}^- = \ClosedLowerEmbed(\vDual{\alpha})
  \rightblackspoon^+ \ClosedDay{I}$. It suffices to prove that
  $\ClosedLowerEmbed(\alpha) \ClosedDay{\bullet}
  \ClosedLowerEmbed(\vDual{\alpha}) \leq \ClosedDay{I}$, which follows
  from the preservation of monoids by $\ClosedLowerEmbed$ and the
  \RuleLabel{Axiom} rule of MAV.

  When $P = Q \vParr R$, $Q \vWith R$, or $Q \vSeq R$, the result
  follows from preservation of the corresponding monoid structure by
  $\ClosedLowerEmbed$. For example,
  $\ClosedLowerEmbed(Q \vParr R) \leq \ClosedLowerEmbed(Q)
  \ClosedDay{\bullet} \ClosedLowerEmbed(R) \leq \sem{Q}^-
  \ClosedDay{\bullet} \sem{R}^- = \sem{Q \vParr R}^-$.

  When $P = Q \vPlus R$, we have
  $\ClosedLowerEmbed(Q \vPlus R) \leq \ClosedLowerEmbed(Q)$ and
  $\ClosedLowerEmbed(Q \vPlus R) \leq \ClosedLowerEmbed(R)$, by the
  \RuleLabel{Left} and \RuleLabel{Right} rules. Therefore,
  $\ClosedLowerEmbed(Q \vPlus R) \leq \ClosedLowerEmbed(Q) \lor
  \ClosedLowerEmbed(R) \leq \sem{Q}^- \lor \sem{R}^- = \sem{Q \vPlus
    R}^-$.

  When $P = Q \vTens R$, we have
  $\sem{Q \vTens R}^- = (\sem{Q}^+ \rightblackspoon \sem{R}^-) \land
  (\sem{R}^+ \rightblackspoon \sem{Q}^-)$. We prove inclusion in the
  left-hand side, the right-hand side is similar. The key property we need to prove is:
  \begin{equation}\label{eq:interaction-leq}
    \ClosedLowerEmbed(Q \vTens R) \ClosedDay{\bullet} (\ClosedLowerEmbed(Q) \rightblackspoon^+ \ClosedDay{I})
    \leq
    \ClosedLowerEmbed(R)
  \end{equation}
  Using the monoidality and monotonicity of $\alpha$, this property is
  implied by the following property in lower sets
  $\LowerSet{\textsc{CFMav}}$:
  \begin{displaymath}
    \LowerEmbed(Q \vTens R) \Day{\bullet} (U(\ClosedLowerEmbed(Q)) \rightblackspoon \Day{I})
    \leq
    \LowerEmbed(R)
  \end{displaymath}
  which follows from the \RuleLabel{Switch} rule of MAV and
  calculation. Using Inequation \ref{eq:interaction-leq}, and property
  (ii) above, we can prove the inequality we need:
  \begin{displaymath}
    \ClosedLowerEmbed(Q \vTens R) \ClosedDay{\bullet} \sem{Q}^+
    \leq
    \ClosedLowerEmbed(Q \vTens R) \ClosedDay{\bullet} (\ClosedLowerEmbed(Q) \rightblackspoon^+ \ClosedDay{I})
    \leq
    \ClosedLowerEmbed(R)
    \leq
    \sem{R}^-
  \end{displaymath}
  Using the residuation property of $\ClosedDay{\bullet}$ we can conclude.
\end{proof}

\begin{theorem}
  If $P \longrightarrow^* I$ in SMAV, then $P \longrightarrow^* I$ in MAV.
\end{theorem}

\begin{proof}
  By Theorem \ref{thm:soundness}, $P \longrightarrow^* I$ in SMAV
  implies $I \leq \sem{P}$. Combined with Proposition
  \ref{prop:embedding-sem}, we have
  $I \leq \lnot \ChuEmbed(\LowerEmbed(P)$. By Definition
  \ref{defn:chu} of the ordering of Chu elements, we have
  $\ClosedLowerEmbed(P) \subseteq \ClosedDay{I}$. Since
  $P \in \ClosedLowerEmbed(P)$, we have $P \in \ClosedDay{I}$, which
  by definition means that $P \longrightarrow^* I$ in MAV.
\end{proof}

### 2024-03-27 doc/sections/mav-semantics.tex
[ agda tex ] Streamline and write up the cut elimination proof

FIXME: cite that combinatorics book.

@@

\begin{definition}\label{defn:mav-interpretation}
  Let $\mathit{At}$ be a set of atomic propositions. Given an
  MAV-algebra $(A, \leq, \otimes, \lhd, I, \lnot)$ and valuation
  $V : \mathit{At} \to A$, define the interpretation of MAV Formulas
  as follows: $\sem{\vUnit} = I$, $\sem{a} = V(a)$,
  $\sem{\vDual a} = \lnot V(a)$,
  $\sem{\vP\vTens\vQ} = \sem{\vP} \otimes \sem{\vQ}$,
  $\sem{\vP\vParr\vQ} = \sem{\vP} \parr \sem{\vQ}$,
  $\sem{\vP\vSeq\vQ} = \sem{\vP};\sem{\vQ}$,
  $\sem{\vP\vWith\vQ} = \sem{\vP} \with \sem{\vQ}$, and
  $\sem{\vP\vPlus\vQ} = \sem{\vP} \oplus \sem{\vQ}$.
\end{definition}

@@

\begin{proposition}\label{prop:day-construction}
  If $(\bullet, i)$ is a pomonoid on $(A, \leq)$, then there is a
  corresponding \emph{Day pomonoid} $(\Day{\bullet}, \Day{i})$ on
  $\LowerSet{A}$ defined as
  $F \Day\bullet G = \{ z \mid z \leq x \bullet y, x \in F, y \in G
    \}$ and $\Day{i} = \eta(i)$. Moreover:
  \begin{enumerate}
    \item $(\Day{\bullet}, \Day{i})$ is left and right residuated: there
          exists a function
          $\Day{\rightblackspoon} : \LowerSet{A} \times \LowerSet{A} \to
            \LowerSet{A}$ such that ... FIXME
    \item If $(\bullet, i)$ is a commutative pomonoid, then so is
          $(\Day{\bullet}, \Day{i})$. In this case, the residuals coincide.
  \end{enumerate}
  \bob{Could also say that $\eta$ preserves the monoids. Does this
    come up later?}
\end{proposition}

\begin{remark}
  This is the Day construction \cite{day} restricted to the case of
  partial orders.
\end{remark}

\begin{remark}
  When $(A, \leq)$ is an MAV-frame, by \ref{prop:day-construction} we
  obtain two pomonoids $(\Day{\parallel}, \Day{I})$ and
  $(\Day{\lhd}, \Day{I})$ on $\LowerSet{A}$. Moreover, the next
  proposition states that the duoidal relationship between these
  monoids is preserved by the Day construction:
\end{remark}

@@

@@

\begin{remark}
  Note that $\alpha$ does not preserve meets.
\end{remark}

\begin{proposition}\label{prop:closed-monoid-distrib}
  For a commutative pomonoid $(\bullet, i)$ on $(A, \leq)$ that
  distributes over $+$
  ($(x + y) \bullet z \leq (x \bullet z) + (y \bullet z)$), then
  $F \ClosedDay{\bullet} G = \alpha(U F \Day{\bullet} U G)$ and
  $\ClosedDay{j} = \alpha(\Day{j})$ define a residuated commutative
  pomonoid on $\ClosedLowerSet{A}$.
\end{proposition}

@@

\begin{proposition}\label{prop:closed-monoid-duoidal}
  For a pomonoid $(\lhd, j)$ on $(A, \leq)$, if this satisfies
  $(w \lhd x) + (y \lhd z) \leq (w + y) \lhd (x + z)$ then the Day
  construction
  $F \Day{\lhd} G = \{ z \mid z \leq x \lhd y, x \in F, y \in G \}$ on
  lower sets is $+$-closed when $F$ and $G$ are. We write
  $F \ClosedDay{\lhd} G$ to indicate when we mean this construction as
  an operation on $+$-closed lower sets. If $j + j \leq j$, then the
  Day unit $\Day{j} = \eta(j)$ is also closed and we write it as
  $\ClosedDay{j} \in \ClosedLowerSet{A}$. Together
  $(\ClosedDay{\lhd}, \ClosedDay{j})$ form a pomonoid on
  $(\ClosedLowerSet{A}, \subseteq)$.
\end{proposition}

@@

\begin{definition}
  Let $k$ be an element of $A$. $\Chu(A, k)$ is the partial order with
  elements pairs $(a^+, a^-)$ such that $a^+ \bullet a^- \leq k$, with
  ordering $(a^+,a^-) \sqsubseteq (b^+, b^-)$ when $a^+ \leq b^+$ and
  $b^- \leq a^-$.
\end{definition}

>>>>

\bob{FIXME: cite that combinatorics book.}

@@

\begin{definition}\label{defn:mav-interpretation}
  Let $\mathit{At}$ be a set of atomic propositions. Given an
  MAV-algebra $(A, \leq, \otimes, \lhd, I, \lnot)$ and valuation
  $V : \mathit{At} \to A$, define the interpretation of MAV Formulas
  as follows: $\sem{\vUnit} = I$, $\sem{\alpha} = V(\alpha)$,
  $\sem{\vDual \alpha} = \lnot V(\alpha)$,
  $\sem{\vP\vTens\vQ} = \sem{\vP} \otimes \sem{\vQ}$,
  $\sem{\vP\vParr\vQ} = \sem{\vP} \parr \sem{\vQ}$,
  $\sem{\vP\vSeq\vQ} = \sem{\vP};\sem{\vQ}$,
  $\sem{\vP\vWith\vQ} = \sem{\vP} \with \sem{\vQ}$, and
  $\sem{\vP\vPlus\vQ} = \sem{\vP} \oplus \sem{\vQ}$.
\end{definition}

@@

\begin{proposition}\label{prop:day-construction}
  If $(\bullet, i)$ is a pomonoid on $(A, \leq)$, then there is a
  corresponding \emph{Day pomonoid} $(\Day{\bullet}, \Day{i})$ on
  $\LowerSet{A}$ defined as
  $F \Day\bullet G = \{ z \mid z \leq x \bullet y, x \in F, y \in G
  \}$ and $\Day{i} = \eta(i)$. Moreover:
  \begin{enumerate}
  \item $(\Day{\bullet}, \Day{i})$ is left and right residuated: there
    exists a function
    $\Day{\rightblackspoon} : \LowerSet{A} \times \LowerSet{A} \to
    \LowerSet{A}$ such that ... \bob{FIXME}
  \item If $(\bullet, i)$ is a commutative pomonoid, then so is
    $(\Day{\bullet}, \Day{i})$. In this case, the residuals coincide.
  \item The embedding preserves the monoid:
    $\eta(x \bullet y) = \eta(x) \Day\bullet \eta(y)$.
  \end{enumerate}
\end{proposition}

\begin{remark}
  This is the Day monoidal product \cite{day} restricted to the case
  of partial orders.
\end{remark}

\begin{remark}
  When $(A, \leq)$ is an MAV-frame, Proposition
  \ref{prop:day-construction} gives us two pomonoids
  $(\Day{\parallel}, \Day{I})$ and $(\Day{\lhd}, \Day{I})$ on
  $\LowerSet{A}$. Moreover, the next proposition states that the
  duoidal relationship between these monoids is preserved by the Day
  construction:
\end{remark}

@@

\begin{definition}
  Define the embedding $\eta^+ : A \to \ClosedLowerSet{A}$ as
  $\eta^+(x) = \alpha(\eta(x))$.
\end{definition}

@@

\begin{proposition}\label{prop:closed-eta-preserve-joins}
  $\eta^+(x + y) \subseteq \eta^+(x) \lor \eta^+(y)$.
\end{proposition}

\begin{remark}
  Proposition \ref{prop:closed-eta-preserve-joins} is the reason for
  requiring $+$-closure. This property will allow us to prove the
  crucial embedding property for all formulas in Section
  \ref{sec:mav-cut-elimination}.

  \bob{Not an equality, I think, because $+$ isn't necessarily a join.}
\end{remark}

\begin{proposition}\label{prop:closed-monoid-distrib}
  For a commutative pomonoid $(\bullet, i)$ on $(A, \leq)$ that
  distributes over $+$
  ($(x + y) \bullet z \leq (x \bullet z) + (y \bullet z)$), then
  $F \ClosedDay{\bullet} G = \alpha(U F \Day{\bullet} U G)$ and
  $\ClosedDay{j} = \alpha(\Day{j})$ define a residuated commutative
  pomonoid on $\ClosedLowerSet{A}$. Moreover,
  $\eta^+(x \bullet y) = \eta^+(x) \ClosedDay{\bullet} \eta^+(y)$.
\end{proposition}

@@

\begin{proposition}\label{prop:closed-monoid-duoidal}
  For a pomonoid $(\lhd, j)$ on $(A, \leq)$, if this satisfies
  $(w \lhd x) + (y \lhd z) \leq (w + y) \lhd (x + z)$ then the Day
  construction
  $F \Day{\lhd} G = \{ z \mid z \leq x \lhd y, x \in F, y \in G \}$ on
  lower sets is $+$-closed when $F$ and $G$ are. We write
  $F \ClosedDay{\lhd} G$ to indicate when we mean this construction as
  an operation on $+$-closed lower sets. If $j + j \leq j$, then the
  Day unit $\Day{j} = \eta(j)$ is also closed and we write it as
  $\ClosedDay{j} \in \ClosedLowerSet{A}$. Together
  $(\ClosedDay{\lhd}, \ClosedDay{j})$ form a pomonoid on
  $(\ClosedLowerSet{A}, \subseteq)$. Moreover,
  $\eta^+(x \lhd y) = \eta^+(x) \ClosedDay{\lhd} \eta^+(y)$.
\end{proposition}

@@

\begin{definition}\label{defn:chu}
  Let $k$ be an element of $A$. $\Chu(A, k)$ is the partial order with
  elements pairs $(a^+, a^-)$ such that $a^+ \bullet a^- \leq k$, with
  ordering $(a^+,a^-) \sqsubseteq (b^+, b^-)$ when $a^+ \leq b^+$ and
  $b^- \leq a^-$.
\end{definition}
